---
title: "Class 6: R Functions Lab"
author: "Nicholas Chiu"
date: 01/25/2024
format: pdf
---
## R Functions

Functions are how we get stuff done. We call functions to do everything useful
in R.

All functions in R have at least 3 things:

- A **name** (we get to pick this)
- One or more **input arguments** (parameters)
- The **body** (lines of code that do the work)

```{r}
#| eval: false

funname <- function(input1, input2) {
  # The body with R code
}
```

Silly function to add 2 numbers

```{r}
x <- 5
y <- 1
x + y
addme <- function(x,y=1) {
  x+y
}
```
```{r}
addme(x,y)
addme(10)
```

## Lab for Today

```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

grade <- function(vec) {
  #vec[is.na(vec)] <- 0
  mean(vec, na.rm = T)
}

grade(student1)
grade(student2)
grade(student3)

# remove lowest score and find the mean
ind <- which.min(student1)
mean(student1[-ind])

```

## Q1: Writing grade()
```{r}
##create function
grade <- function(x) {
  #change NA to 0
  x[is.na(x)] <- 0
  
  #calculate mean
  mean(x[-which.min(x)])
}

grade(student3)
```

```{r myfun}
# Name: grades
# Parameters: CSV file containing a vector of student grades
# Output: 
# grade <- function(url) {
#   # Read in CSV
#   gradebook <- read.csv(url, row.names = 1)
#   x[is.na(x)] <- 0
#   mean(x[-which.min(x)])
#   
# }


url <- "https://tinyurl.com/gradeinput/"
gradebook <- read.csv(url, row.names = 1)

```


```{r}
results <- apply(gradebook,1,grade)
gradebook$avg <- apply(gradebook,1,grade)
head(gradebook)
```

## Q2: Top Scoring Student
```{r}
which.max(results)
```
Student 18 scored the highest.

## Q3: Toughest HW
```{r}
# Lowest score based on mean
which.min(apply(gradebook, 2, mean, na.rm=T))

# Lowest score based on sum
which.min(results2 <- apply(gradebook, 2, sum, na.rm=T))
```
HW2 had the lowest scores overall. 

## Q4: Most predictive

```{r}
mask <- gradebook
mask[is.na(mask)] <- 0
#mask

cor(mask$hw1, results)
cor(mask$hw5, results)

# Use 'apply()' function to run 'cor()' cover the whole course (ie. masked gradebook)
correlation <- apply(mask, 2, cor, results)
which.max(correlation)
```
HW6 had the most predictive overall score.
